---
title: "Class 11"
author: "Erica Sanchez (A15787505)"
date: "May 12, 2024"
format: pdf
---
```{r}
mxl <- read.csv("SampleGenotypes.csv")
head(mxl)
```
# Section 1 - proportion of (G/G) in a population

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
(table(mxl$Genotype..forward.strand.) / nrow(mxl)) *100
```

# Section 4 - Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
# read the dataset
filegp <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt", header = T)
```

```{r}
summary(filegp)
```
```{r}
str(filegp)
# calculate the median expression level for each genotype
medians <- tapply(filegp$exp, filegp$geno, median)
print(medians)
# calculate the sample size for each genotype
sample_sizes <- tapply(filegp$geno, filegp$geno, length)
print(sample_sizes)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
boxplot(exp ~ geno, data = filegp, xlab = "geno", ylab = "exp", main = "exp of ORMDL3 by geno")
```

The SNP affects the expression of ORMDL3 because the boxplot shows the median expression levels of A/A are higher than the median expression levels of G/G.